---
title: "Report_test"
author: "toprak"
date: "`r Sys.Date()`"
output: html_document
---


## Original data

The data comes from randomized trial,
[here are details](https://github.com/toprakd/group5/blob/main/DATA/codebook_updated)


### Reading data 
```{r include=FALSE}
library(tidyverse)
# install.packages("tidyverse")
library(here)


#read data
exam_nontidy <- read_delim(here("DATA", "exam_nontidy.txt"))
```



```{r}
exam_nontidy
```

##Tidying the data

to tidy the data, we had to:

- remove duplicated rows
- rename coloumns that contain spaces or start with numbers.
- separate subject and pregnancy_num-age coloumns into two
- create to new coloums called glucose_mg_dl and dbp_mm_hg because the .value coloumn was including values from both features.
- remove unnecessary "cholesterol" and "sibling" columns 
- add a numeric column showing insulin in units pmol/L (1 microiu_ml = 6.945 pmol_l)
    

        - some columns can include values from various features/measurements
?
```{r} 



#Changing the names of coloumns
exam_nontidy <- 
  exam_nontidy %>% 
  rename(insulin_microiu_ml = `insulin microiu ml`,
         diabetes_5_year = `5y diabetes`,
         cholesterol = choleste)


#Separate variable subject into hospital and ID:
df <- separate(exam_nontidy, subject, sep ="-",
               into = c("hospital", "ID"))


#Separate variable pregnancy_num-age into pregnancy_num and age:
df1 <- separate(df, "pregnancy_num-age", sep ="-",
                into = c("preg_num", "age"))


#remove duplicates and pivot wider (for the measured variable and the .value coloumn)
exam_tidy <- df1 %>%
  distinct() %>%   #remove duplicates 
  pivot_wider(names_from = `measured variable`, 
              values_from = .value) #.value include values from 2 features/measurements

```



## Wrangling ----
Read and join the additional dataset to your main dataset 
```{r}
join_data <- read.delim(here("DATA/exam_joindata.txt"))
join_data <- join_data %>%
  rename(ID = id)
```
Merge steps
```{r}
tidy_data <- tidy_data %>% 
  select(- cholesterol, - sibling)  %>% 
  mutate(glucose_mg_dl_categoric = ifelse(glucose_mg_dl >120, "High", "Low"),
         insulin_pmol_L = insulin_microiu_ml * 6.945,
         diabetes_5_year_binary= ifelse(diabetes_5_year== "neg",0,1),
         multiplication_age_and_pregnancy_num= preg_num*age) %>% 
  select(ID, hospital, age, everything()) %>% 
  arrange(ID) %>%
  full_join(join_data, join_by("ID")) 

tidy_data  %>% 
  group_by(hospital) %>% 
  summarise( min(preg_num), max(preg_num ), mean(preg_num),sd(preg_num ))

```

Summary and overview of our data 
```{r}
summary(tidy_data) 
glimpse(tidy_data)
skimr::skim(tidy_data)
```

Explore and comment on missing data 
```{r}
naniar::gg_miss_var(tidy_data)
```

Stratefying data 
```{r}
tidy_data %>%
  group_by(hospital) %>%
  summarise(min(preg_num), max(preg_num), mean(preg_num), sd(preg_num))
```

Diabetes 5 year pos 
```{r}
tidy_data %>%
  filter(diabetes_5_year_binary == 1) %>%
  group_by(hospital) %>%
  summarise(min(preg_num), max(preg_num), mean(preg_num), sd(preg_num))
```

Only for persons recruited in Hosp1 
```{r}
tidy_data  %>% 
  group_by(hospital) %>% 
  summarise( min(preg_num), max(preg_num ), mean(preg_num),sd(preg_num )) %>%
  filter(hospital == "Hosp1")
```

Only for persons with `pregnanyc_num` more than 2 
```{r}
tidy_data  %>% 
  filter(preg_num>2)%>% 
  group_by(hospital) %>% 
  summarise( min(preg_num), max(preg_num), mean(preg_num),sd(preg_num ))
```

`glucose_mg_dl` lower than 120 and `dbp_mm_hg` greater than 60
```{r}
tidy_data  %>% 
  filter(glucose_mg_dl < 120, dbp_mm_hg > 60) %>% 
  group_by(hospital) %>% 
  summarise( min(preg_num), max(preg_num ), mean(preg_num),sd(preg_num ))
```

Use two categorical columns in your dataset to create a table (hint: ?count)
```{r}
tidy_data %>%
  group_by(hospital, diabetes_5_year) %>%
  summarise(count = n())
```


## Plots


## Statistical analysis ----
```{r}
exam_tidy %>% 
  kruskal.test(diabetes_5_year~pedigree, data = .) %>%
  broom::tidy()
```
The effect of pedigree on diabetes outcome was not statistically significant so diabetes does not seems to be effected by pedigree  

```{r}
exam_tidy %>% 
  kruskal.test(diabetes_5_year~bmi, data = .) %>%
  broom::tidy()
```
The effect of BMI on diabetes was statistically significant so diabetes seems to be effected by BMI

```{r}
exam_tidy %>% 
  kruskal.test(diabetes_5_year~glucose_mg_dl, data = .) %>%
  broom::tidy()
```
The effect of glucose_mg_dl on diabetes was statistically significant so so diabetes seems to be effected by glucose levels 

```{r}
exam_tidy %>% 
  kruskal.test(diabetes_5_year~hospital, data = .) %>%
  broom::tidy()
```
The effect of hospital on diabetes was not statistically significant so diabetes seems to be effected by hospital


```{r}
sessionInfo()
```





