---
title: "Final report group 5"
author: "Toprak, Siren, Cathrine"
date: "`r Sys.Date()`"
output: html_document
---


## Original data

The data comes from Diabetes Prediction Dataset from the Pima Indian Tribe and the NIDDK,
[here are details](https://github.com/toprakd/group5/blob/main/DATA/codebook_updated).


### READING DATA 
```{r include=FALSE}
library(tidyverse)
# install.packages("tidyverse")
library(here)

#read data
exam_nontidy <- read_delim(here("DATA", "exam_nontidy.txt"))
```


```{r}
exam_nontidy
```

## TIDYING THE DATA
To tidy the data, we had to:

- remove duplicated rows
- rename coloumns that contain spaces or start with numbers
- separate subject and "pregnancy_num-age" coloumns into two
- create to new coloums called "glucose_mg_dl" and "dbp_mm_hg" because the ".value" coloumn was including values from both features

```{r} 
#Changing the names of coloumns:
exam_nontidy <- 
  exam_nontidy %>% 
  rename(insulin_microiu_ml = `insulin microiu ml`,
         diabetes_5_year = `5y diabetes`,
         cholesterol = choleste)

#Separate variable subject into "hospital" and "ID":
df <- separate(exam_nontidy, subject, sep ="-",
               into = c("hospital", "ID"))

#Separate variable "pregnancy_num-ag"e into "pregnancy_num" and "age":
df1 <- separate(df, "pregnancy_num-age", sep ="-",
                into = c("preg_num", "age"))

#remove duplicates and pivot wider (for the measured variable and the ".value" coloumn):
tidy_data <- df1 %>%
  distinct() %>%    
  pivot_wider(names_from = `measured variable`, 
              values_from = .value) 
```


## WRANGLING 

We did the following steps to further tidy our data:

- remove unnecessary "cholesterol" and "sibling" columns 
- correct the variable types 
- adding a numeric column showing insulin in units pmol/L (1 microiu_ml = 6.945 pmol_l)
- adding a column showing `diabetes_5_year` as 0/1
- adding a numeric column showing multiplication of `age` and `pregnancy_num` for each person
- setting the order of columns as: `ID, hospital, age` and other columns
- arranged ID column of the dataset in order of increasing number 


Then we merged our dataset with an additional dataset (exam_joindata). [Here are details for the additional data](https://github.com/toprakd/group5/blob/main/DATA/exam_joindata.txt.).

```{r include=FALSE}
join_data <- read.delim(here("DATA/exam_joindata.txt"))
join_data <- join_data %>%
  rename(ID = id)
```

```{r}
tidy_data <- tidy_data %>% 
  select(- cholesterol, - sibling)  %>% 
  mutate(across(ID:age, as.numeric)) %>%
  mutate(glucose_mg_dl_categoric = ifelse(glucose_mg_dl >120, "High", "Low"),
         insulin_pmol_L = insulin_microiu_ml * 6.945, 
         diabetes_5_year_binary= ifelse(diabetes_5_year== "neg",0,1),
         multiplication_age_and_pregnancy_num= preg_num*age) %>% 
  select(ID, hospital, age, everything()) %>% 
  arrange(ID) %>%
  full_join(join_data, join_by("ID")) 

```

Summary and overview of our data:
```{r echo=FALSE}
summary(tidy_data)
```
- Triceps_mm, insulin_microiu_ml, bmi, diabetes_5_year, dbp_mm_hg, insulin_pmol_L and antibody had some NA vaules.


Exploration of missing data 

```{r echo=FALSE}
naniar::gg_miss_var(tidy_data)
```

### Stratification

Then we did the following stratification to explore our data:

- Pregnancy numbers in different hospitals

```{r echo=FALSE}
tidy_data %>%
  group_by(hospital) %>%
  summarise(min(preg_num), max(preg_num), mean(preg_num), sd(preg_num))
```

- Pregnancy numbers in different hospitals filtered by "diabetes_5_year" positive patiens 

```{r echo=FALSE}
tidy_data %>%
  filter(diabetes_5_year_binary == 1) %>%
  group_by(hospital) %>%
  summarise(min(preg_num), max(preg_num), mean(preg_num), sd(preg_num))
```

- Pregnancy numbers in Hospital 1

```{r echo=FALSE}
tidy_data  %>% 
  group_by(hospital) %>% 
  summarise( min(preg_num), max(preg_num ), mean(preg_num),sd(preg_num )) %>%
  filter(hospital == "Hosp1")
```

- Only for persons with `pregnanyc_num` more than 2 

```{r echo=FALSE}
tidy_data  %>% 
  filter(preg_num>2)%>% 
  group_by(hospital) %>% 
  summarise( min(preg_num), max(preg_num), mean(preg_num),sd(preg_num ))
```

- For `glucose_mg_dl` lower than 120 and `dbp_mm_hg` greater than 60

```{r echo=FALSE}
tidy_data  %>% 
  filter(glucose_mg_dl < 120, dbp_mm_hg > 60) %>% 
  group_by(hospital) %>% 
  summarise( min(preg_num), max(preg_num ), mean(preg_num),sd(preg_num ))
```

- Patient numbers according to different hospitals and diabetes status
```{r echo=FALSE}
tidy_data %>%
  group_by(hospital, diabetes_5_year) %>%
  summarise(count = n())
```


## PLOTS 
We made plots to check if there was any correlated measurements:

```{r echo=FALSE}
# Checking for correlated measurements, computing correlation matrix and plotting:

library(ggcorrplot)
numeric_data <- tidy_data %>%
  select(where(is.numeric))

corr <- cor(numeric_data, use = "complete.obs")

ggcorrplot(
  corr,
  method = "square",           
  type = "full",                #
  ggtheme = ggplot2::theme_minimal(),
  title = "Correlation Matrix",
  show.legend = TRUE,
  legend.title = "Corr",
  show.diag = TRUE,             
  colors = c("blue", "white", "red"),
  outline.color = "gray",
  hc.order = FALSE,             
  lab = TRUE,                   
  lab_col = "black",
  lab_size = 4,
  tl.cex = 12,                  
  tl.col = "black",
  tl.srt = 45,                  
  digits = 2                   
)
```

- In the correlation matrix we mostly saw positive correlations. Especially, "glucose_mg_dl" and "insulin_microiu_ml" correlated, and bmi and "triceps_mm" were correlated.


Created plot to examine if the level of glucose and insulin depend on each other:

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(data = tidy_data) +
  aes(
    x = glucose_mg_dl,
    y = insulin_microiu_ml
  ) + 
  geom_point() +
  geom_smooth(method = "lm")

```

- The glucose and insulin level seems to be dependent on each other.


Examined if the level of glucose and insulin depend on each other when stratifying by outcome (`diabetes_5_year`):

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(data = tidy_data) +
  aes(
    x = glucose_mg_dl,
    y = insulin_microiu_ml
  ) +
  geom_point(
    aes(color = diabetes_5_year,
        shape = diabetes_5_year),
    size = 2 
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    aes(color = diabetes_5_year)
  )
```

- The level of glucose and insulin were correlated in both 5 year diabetes positive and negative patients, but the correlation seemed stronger in diabetes negative patients. 


Examine the correlation between glucose and blood pressure: 

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(data = tidy_data) +
  aes(
    x = glucose_mg_dl,
    y = dbp_mm_hg
  ) +
  geom_point() +
  geom_smooth(method = "lm")
```

- There does not seem to be strong correlation between glucose and blood pressure.


Examine the relation between the `BMI` and `triceps_mm` values:

```{r echo=FALSE, warning=FALSE, message=FALSE}

ggplot(data = tidy_data) +
  aes(
    x = bmi,
    y = triceps_mm
  ) +
  geom_point() +
  geom_smooth(method = "lm")
```

- There was a strong relationship between BMI and triceps_mm.


Examine the blood pressure distribution in different BMI categories:

```{r echo=FALSE, warning=FALSE, message=FALSE}

# Creating BMI categories:
tidy_data <- tidy_data %>%
  mutate(
    bmi_category = case_when(
      bmi < 25 ~ "healthy_weight",
      bmi >= 25 & bmi < 30 ~ "over_weight",
      bmi >= 30 ~ "obesity"
    ),
    # Order of levels:
    bmi_category = factor(bmi_category, levels = c("healthy_weight", "over_weight", "obesity"))
  )

# Plotting:
ggplot(data = tidy_data) +
  aes(
    x = bmi_category,
    y = dbp_mm_hg
  ) +
  geom_boxplot()
```

- Blood pressure distributions tend to increase with higher obesity levels based on BMI categories. In patients with unknown BMI values, the pattern closely resembles that of the obesity group.


## STATISTICAL ANALYSIS 

We analysed the datasets to check if diabetes outcome were dependent on the following variables:

Pedigree
```{r echo=FALSE}
tidy_data %>% 
  kruskal.test(diabetes_5_year~pedigree, data = .) %>%
  broom::tidy()
```
- The effect of pedigree on diabetes outcome was not statistically significant so diabetes does not seems to be effected by pedigree  

BMI
```{r echo=FALSE}
tidy_data %>% 
  kruskal.test(diabetes_5_year~bmi, data = .) %>%
  broom::tidy()
```
- The effect of BMI on diabetes was statistically significant so diabetes seems to be effected by BMI

Glucose
```{r echo=FALSE}
tidy_data %>% 
  kruskal.test(diabetes_5_year~glucose_mg_dl, data = .) %>%
  broom::tidy()
```
- The effect of glucose_mg_dl on diabetes was statistically significant so so diabetes seems to be effected by glucose levels 

Hospital
```{r echo=FALSE}
tidy_data %>% 
  kruskal.test(diabetes_5_year~hospital, data = .) %>%
  broom::tidy()
```
- The effect of hospital on diabetes was not statistically significant so diabetes seems to be effected by hospital







